### 2020-05-16 Action-letter.Rmd
Final changes for print version

---
title       : "Quantitative involvement moderates longitudinal associations between mothers' and fathers' qualitative parenting and children's emotional adjustment"
authors     : "Caspar J. van Lissa & Renske Keizer"
journal     : "Developmental Psychology"
manuscript  : "DEV-2019-2824"
bibliography: ["DSEM_2017.bib"]  
class       : "draft"

output      : papaja::revision_letter_pdf
---

@@

\subsection*{EDITOR'S NOTE: Please make sure your revision is concise. Our guidelines for length of manuscripts are 35 pages inclusive of everything except tables and figures. Because your manuscript reports on a longitudinal study, we could allow a few extra pages, but please keep close to this guideline.}

@@

With regard to the Editorial Office's request that we upload a Word or RTF file: In line with the Submission Guidelines on the Developmental Psychology website, we have provided \latex as a `zip`-file, with accompanying `.pdf`. We do not have a Word version of this manuscript available. We could manually convert this to Word, but the typographical advantages of using Latex would then be lost. Please let us know how to proceed.

@@

\WorkInProgress

@@

> Bayes is a full-information estimator, which means that all available information is used without imputing missing values, and consistent estimation is guaranteed under the assumption that data are missing at random. The missing data handling is likelihood based and guarantees consistent estimation as long as the missing data is missing at random (MAR).

@@

We appreciate the suggestion, and have incorporated this Figure, see below:

```{r figure1, out.width="100%", fig.cap="T-1: previous measurement occasion. ri: random intercept. s: random slope. M: Mother's qualitative involvement. E: Emotional adjustment. F: Father's qualitative involvement. inv: quantitative involvement. Thus, sME represents the random slope of Mothers' qualitative involvement on emotional adjustment."}
knitr::include_graphics("Figure1.png")
```

@@

> In the area of parenting and emotional adjustment development, we are aware of only one prior study that controlled for stable between-family differences, and similarly reported mostly child effects and between-family level correlations in a sample of adolescents [@vanlissaRoleFathersMothers2018a]. The authors explained their findings by proposing that child effects might be more common in adolescence than in childhood, because parent-adolescent relationships are more egalitarian at that age. They argued that the between-family level correlations might reflect the consolidated influence of within-family parent effects that occurred at an earlier age. The fact that the present study similarly found a preponderance of child effects and between-family level correlations in a sample of younger children, age 6-12, suggests that __either parent effects occurred at an even younger age__, or that something else is at play.

@@

\WorkInProgress

@@

\WorkInProgress

@@

\WorkInProgress

@@

We appreciate the suggestion; indeed, the idea that both quantitative and qualitative involvement are important (and might interact) is most important. The rough division along disciplinary lines is interesting, but not central to the argument. We have therefore edited our argumentation to focus on the importance of jointly investigating quantitative and qualitative involvement, and removed all references to disciplinary differences.

@@

\WorkInProgress

@@

> We therefore hypothesize that parents will display more warmth, consistency, and inductive reasoning towards girls than boys, and more anger towards boys than girls, $H_{7a-d}$. Furthermore, we predict stronger associations between qualitative parenting and emotional adjustment between children and their same-sex parent, $H_{8a-d}$.
*[...]*
Consequently, we have no hypothesis regarding sex differences in children's average levels of emotional adjustment.

@@

> Bayes is a full-information estimator, which means that all available information is used without imputing missing values, and consistent estimation is guaranteed under the assumption that data are missing at random.

@@

> Between-family differences in the interplay between qualitative parenting behaviors and emotional adjustment are estimated as random effects, and we can include between-family level moderators as predictors of these random effects. In this way, we included total joint childcare, the childcare distribution, __and child sex__ as moderators.

@@

> We examined differences between fathers and mothers in parent- and child effects the same way [by testing the significance of the difference between corresponding mother- and father effects (see Table 2 and Supplementary Table S5)]. All effects that were significant for one parent but not significant for the other parent had significant difference tests, except for the child effect of emotional adjustment on parental anger, which did not differ significantly between parents.

@@

> Fathers' qualitative involvement is associated with emotional adjustment, even after accounting for mothers' qualitative involvement [@flouriFathersBehaviorsChildren2010]. Although some research suggests that both parents influence children similarly [@lambDevelopmentSignificanceFatherChild2004; @faganShouldResearchersConceptualize2014], there is also evidence for unique contributions of mothers and fathers. Some have theorized that mothers and fathers might fulfill somewhat specialized roles [@bogelsDoesFatherKnow2011]. In line with the theoretical notions of "attachment" and "activation" relationships, mothers tend to encourage and respond supportively to emotional expression to a greater extent than fathers, whereas fathers structure and sanction emotion expressions more than mothers [@cassanoInfluenceGenderParental2007; @eisenbergParentalReactionsChildren1999]. Moreover, for adolescents' emotional adjustment, maternal warmth was found to be a more salient socialization mechanism than demandingness, whereas this pattern was reversed for fathers [@vanlissaRoleFathersMothers2018a]. Similar differentiated associations were also found for younger children's psychological adjustment [@mcdowellDifferencesMothersFathers2003]. Finally, there is some evidence that emotional adjustment is more negatively impacted by mothers' negative parenting, compared to fathers [@changHarshParentingRelation2003]; perhaps because negative parenting is incongruous with the notion of the maternal "attachment" relationship. We might therefore hypothesize that emotional adjustment would be more strongly associated with maternal warmth and anger, and with paternal demanding behaviors, such as consistency and inductive reasoning, $H_{3}$.

@@

> Negative parenting, which we operationalized as parental anger, is an additional qualitative parenting dimension that is negatively linked with children's emotional adjustment [@changHarshParentingRelation2003; @eisenbergParentalReactionsChildren1999]. 
Parental anger expressions negatively impact children's emotional adjustment [@scaramellaParentingChildrenDistress2008], because parental anger is upsetting, and because anger reflects poor parental emotion regulation, which serves as a model for children [@changHarshParentingRelation2003].

@@

> [...] bi-directional effects are not routinely investigated in the childhood literature [@flouriFathersBehaviorsChildren2010]. Consequently, most of the research cited in support of hypotheses $H_{1a-d}$ is cross-sectional, and could be taken as support for child effects as well as parent effects. We therefore hypothesize that child effects are similar in direction, and that warmth, consistency, and inductive reasoning positively predict, and anger negatively predicts, emotional adjustment, $H_{2a-d}$. This is supported by studies that investigated child effects specifically, and found that children's maladjustment elicits less parental warmth, and more negative parenting [@gadeyneLongitudinalRelationsParenting2004]. Bi-directional positive links between parental consistency and emotional adjustment have also been reported [@lenguaBidirectionalAssociationsTemperament2005]. 
The adolescent literature provides additional evidence that emotional adjustment positively predicts parental support [@otterpohlCrossLaggedRelationsParenting2015; @vandergiessenPerceivedAutonomySupport2014], and negatively predicts negative parenting [@wijsbroekDirectionEffectsPerceived2011]. Thus, in a way, parents respond positively to emotional adjustment, and negatively to maladjustment.

@@

> we are aware of only one prior study that controlled for stable between-family differences, and similarly reported mostly child effects and between-family level correlations in a sample of adolescents [@vanlissaRoleFathersMothers2018a]. The authors explained their findings by proposing that child effects might be more common in adolescence than in childhood, because parent-adolescent relationships are more egalitarian at that age. They argued that the between-family level correlations might reflect the consolidated influence of within-family parent effects that occurred at an earlier age. The fact that the present study similarly found a preponderance of child effects and between-family level correlations in a sample of younger children, age 6-12, suggests that either parent effects occurred at an even younger age, or that something else is at play.

@@

> We found only limited evidence for within-family predictive effects, most of which were child effects on parents. Instead, we mostly found evidence for between-family level correlations. We are not the first to find that parent effects are less prevalent once stable between-family differences are accounted for [@keijsersParentalMonitoringAdolescent2016]. In the area of parenting and emotional adjustment development, we are aware of only one prior study that controlled for stable between-family differences, and similarly reported mostly child effects and between-family level correlations in a sample of adolescents [@vanlissaRoleFathersMothers2018a]. The authors explained their findings by proposing that child effects might be more common in adolescence than in childhood, because parent-adolescent relationships are more egalitarian at that age. They argued that the between-family level correlations might reflect the consolidated influence of within-family parent effects that occurred at an earlier age. The fact that the present study similarly found a preponderance of child effects and between-family level correlations in a sample of younger children, age 6-12, suggests that either parent effects occurred at an even younger age, or that something else is at play.  
  As similar patterns continue to emerge, it might be worth reflecting on their implications for developmental theory. Our findings are less congruent with a fully bidirectional conceptualization of reciprocal influence, and more consistent with a model where parents "set the stage" for children's development through their stable differences from other parents, and continue to adjust to children's changing needs over time. Our findings serve as a reminder that we should be critical of causal interpretations of research that does not control for stable between-family differences, not only in the area of emotional adjustment [e.g., @brandPerceivedParentalInconsistency1990], but in developmental research in general. We should reconsider the question: To what extent are parents shaping children's development, and to what extent are they adapting to it?  
  Future research might also devote greater attention to between-family level correlations as an outcome of interest. These correlations can reflect the consolidated effects of parenting effects that occur at an early age [@vanlissaRoleFathersMothers2018a], or alternatively, they can reflect the influence of third variables that impact both constructs. We should seek to understand why between-family level correlations are so prominent; for example, by comparing their strength in samples with children of different ages, including very young children, or by examining the influence of more sociological factors. For example, economic hardship undermines parents' capacity for consistent parenting, as well as children's socio-emotional adjustment [@mcloydImpactEconomicHardship1990].

@@

> We established direction of effects by testing the significance of the difference between corresponding parent- and child effects (see Table 2 and Supplementary Table S5). All effects that were significant in one direction, and non-significant in the opposite direction, had significant difference tests. All effects that were significant (or non-significant) in both directions had non-significant difference tests.

@@

> To address our research questions, we used multilevel, autoregressive structural equation modeling [@asparouhovDynamicStructuralEquation2018]. In this approach, the variance in qualitative parenting behaviors and emotional adjustment is partialized into stable between-individual differences (the mean level for that individual across all measurement occasions), and within-individual fluctuations (deviations from this person-specific mean). This has several advantages: First, controlling for stable between-family differences allows us to estimate the direction of within-family effects more accurately [@hamakerCritiqueCrosslaggedPanel2015]. Any stable factors which might bias cross-lagged effects - such as the influence of SES or neighborhood effects - are accounted for at the between level, and reflected in the correlations between the random intercepts. Secondly, this approach enables us to directly test our hypothesis that between-family differences in quantitative involvement moderate the interplay between qualitative involvement and emotional adjustment: Between-family differences in the interplay between qualitative parenting behaviors and emotional adjustment are estimated as random effects, and we can include between-family level moderators as predictors of these random effects. In this way, we included joint and relative quantitative involvement, and child sex as moderators. Figure \@ref(fig:figure1) shows a graphical representation of the resulting model. We conducted separate analyses for the four qualitative parenting dimensions, because a model including all four variables at once was too complex to estimate.

@@

> Counter to our hypothesis $H_{6a}$, we found that greater joint involvement was negatively associated with emotional adjustment. Thus, the children in families with highly involved parents generally displayed poorer emotional adjustment. However, hypothesis $H_{6b}$, that children's emotional adjustment would be lower in families where mothers were relatively more involved than fathers, was supported.

@@

\section*{References}

\setlength{\leftskip}{0.4in}
\setlength{\parindent}{-0.4in}

>>>>

---
title       : "Quantitative involvement moderates longitudinal associations between mothers' and fathers' qualitative parenting and children's emotional adjustment"
authors     : "Caspar J. van Lissa & Renske Keizer"
journal     : "Developmental Psychology"
manuscript  : "DEV-2019-2824"
bibliography: ["DSEM_2017.bib"]  
class       : "final"
header-includes:
   - \usepackage{graphicx}
output      : papaja::revision_letter_pdf
---

@@

@@

You requested that we make sure the revision is as concise as possible, and offered that you could allow a few extra pages given the complexity of the study. We have made extensive efforts to shorten the manuscript. We count 10.510 words including references, which is 10 words more than the length indicated for a  larger research report in the [Submission guidelines](https://www.apa.org/pubs/journals/dev/?tab=4). We therefore hope that you can allow us this extra space.

With regard to the Editorial Office's request that we upload a Word or RTF file: In line with the [Submission guidelines](https://www.apa.org/pubs/journals/dev/?tab=4), we have provided \LaTeX\  as a `zip`-file, with accompanying `.pdf`. We do not have a Word version of this manuscript available. We tried to convert this PDF to Word, but the result was completely unacceptable. Please let us know if further action is required. Relatedly, the `zip` contains an EPS file for Figure 1, to ensure that the published version will be rendered in high quality (we have previously published a PNG figure in *Developmental Psychology*, which was illegible in print).

@@

We have done as requested; the first two paragraphs now introduce the two major goals of the paper, and follow the three-point argumentation structure you suggested. We have marked this argumentation structure in the paragraphs below:

> [...] **(1)** Ample research has demonstrated associations between children's emotional adjustment and parenting; both in terms of quantitative involvement (the amount of time parents spend caring for their child), and in terms of qualitative parenting behaviors (how parents behave towards their child). **(2)** Nevertheless, the existing literature has two major shortcomings. Firstly, although it is widely acknowledged that quantitative and qualitative involvement are both relevant to child development [@faganShouldResearchersConceptualize2014; @pleckPaternalInvolvementRevised2004], these dimensions are rarely jointly investigated, let alone in interaction [@craigDoesFatherCare2006; @faganShouldResearchersConceptualize2014]. Therefore, we currently do not know how the interplay between quantitative and qualitative involvement relates to emotional development. For example, children’s emotional adjustment might be differently impacted by anger coming from a highly involved parent, than from a generally absent parent. Conversely, children’s poor emotional adjustment might elicit warm responses from hands-on parents, who are aware of the reasons for their children’s emotional troubles, than from out-of-touch parents, who might perceive maladjustment as a problem behavior that should be disciplined with anger or inductive reasoning. **(3)** The present study overcomes this limitation by investigating both qualitative and quantitative involvement, and the interplay between them, in relation to children's emotional adjustment.  
  Secondly, **(1)** although emotional adjustment is one particular area of development where fathers have been theorized to play a unique role [@paquetteTheorizingFatherchildRelationship2004], fathers are still not consistently included in research [@faganShouldResearchersConceptualize2014]. **(2)** Consequently, prior research has not sufficiently addressed mothers' and fathers' joint and unique contributions to emotional adjustment development. **(3)** The present study overcomes this limitation by incorporating both parents. We use sophisticated methods to examine how joint and relative quantitative involvement at age 6 moderate the longitudinal interplay between mothers' and fathers' qualitative parenting and children's emotional adjustment, from the age of 6 to 12 years. We thus obtain a more comprehensive and nuanced understanding of how fathers' and mothers' involvement relate to children's emotional adjustment development.

@@

> Bayes is a full-information estimator, which means that all available information is used without imputing missing values, and consistent estimation is guaranteed under the assumption that data are missing-at-random (a non-parameteric MCAR test resulted in $p > .05$).

@@

We appreciate the suggestion, and have incorporated this Figure in the revised manuscript (Figure 1).

@@

> In the area of parenting and emotional adjustment development, we are aware of only one prior study that controlled for stable between-family differences, and similarly reported mostly child effects and between-family level correlations in a sample of adolescents [@vanlissaRoleFathersMothers2019]. The authors explained their findings by proposing that child effects might be more common in adolescence than in childhood, because parent-adolescent relationships are more egalitarian at that age. They argued that the between-family level correlations might reflect the consolidated influence of within-family parent effects that occurred at an earlier age. The fact that the present study similarly found a preponderance of child effects and between-family level correlations in a sample of younger children, age 6-12, suggests that __either parent effects occurred at an even younger age__, or that something else is at play.

@@

First, to emphasize that we are studying emotional adjustment in children, we now begin the opening paragraph with *Children's emotional adjustment is defined as...*, and we conclude the second opening paragraph with *We [...] examine [...] mothers' and fathers' qualitative parenting and children's emotional adjustment, from the age of 6 to 12 years.* The developmental age is also mentioned in the paper Title, and in the Abstract.

Second, we have rewritten the first two paragraphs of the *Introduction*, following a structure suggested by the Editor. The Editor wrote: *"Readers want to know (1) what is already known about the association between the parent-child relationship and children's adjustment? (2) what is not known about the association between the parent-child relationship and children's adjustment? (3) how will your study fill that gap in knowledge?"* We have marked this argumentation structure in the paragraphs below:

> Children's emotional adjustment is defined as the absence of internalizing symptoms and emotional dysregulation [@goodmanPsychometricPropertiesStrengths2001]. It is integral to social competence, academic achievement, and mental health [@grossHandbookEmotionRegulation2013]. **(1)** Ample research has demonstrated associations between children's emotional adjustment and parenting; both in terms of quantitative involvement (the amount of time parents spend caring for their child), and in terms of qualitative parenting behaviors (how parents behave towards their child). **(2)** Nevertheless, the existing literature has two major shortcomings. Firstly, although it is widely acknowledged that quantitative and qualitative involvement are both relevant to child development [@faganShouldResearchersConceptualize2014; @pleckPaternalInvolvementRevised2004], these dimensions are rarely jointly investigated, let alone in interaction [@craigDoesFatherCare2006; @faganShouldResearchersConceptualize2014]. Therefore, we currently do not know how the interplay between quantitative and qualitative involvement relates to emotional development. For example, children’s emotional adjustment might be differently impacted by anger coming from a highly involved parent, than from a generally absent parent. Conversely, children’s poor emotional adjustment might elicit warm responses from hands-on parents, who are aware of the reasons for their children’s emotional troubles, than from out-of-touch parents, who might perceive maladjustment as a problem behavior that should be disciplined with anger or inductive reasoning. **(3)** The present study overcomes this limitation by investigating both qualitative and quantitative involvement, and the interplay between them, in relation to children's emotional adjustment.  
  Secondly, **(1)** although emotional adjustment is one particular area of development where fathers have been theorized to play a unique role [@paquetteTheorizingFatherchildRelationship2004], fathers are still not consistently included in research [@faganShouldResearchersConceptualize2014]. **(2)** Consequently, prior research has not sufficiently addressed mothers' and fathers' joint and unique contributions to emotional adjustment development. **(3)** The present study overcomes this limitation by incorporating both parents. We use sophisticated methods to examine how joint and relative quantitative involvement at age 6 moderate the longitudinal interplay between mothers' and fathers' qualitative parenting and children's emotional adjustment, from the age of 6 to 12 years. We thus obtain a more comprehensive and nuanced understanding of how fathers' and mothers' involvement relate to children's emotional adjustment development.

@@

We have done as requested in the re-written section on *Differences between mothers and fathers*:

> The notion that fathers should share child-rearing responsibilities with mothers has become widely accepted in the Western world [@pleckFatherhoodIdealsUnited1997]. Consequently, researchers have come to study qualitative involvement along the same dimensions for mothers and fathers [@maccobySocializationContextFamily1983; @pleckPaternalInvolvementRevised2004]. Fathers' qualitative involvement is associated with emotional adjustment, even after accounting for mothers [@flouriFathersBehaviorsChildren2010]. Some research suggests that both parents influence children similarly, but that effects might be stronger for mothers [@lambDevelopmentSignificanceFatherChild2004; @faganShouldResearchersConceptualize2014]. Others have argued that mothers and fathers might contribute through somewhat different mechanisms. Attachment relationships are grounded in a balance between warmth and security, and exploration [@grossmannDeveloppementAttachementAdaptation1998]. Some have proposed that warmth is more central to mother-child attachment relationships, whereas father-child relationships can be construed as "activation relationships", with greater emphasis on exploration [@paquetteFatherChildActivationRelationship2013]. By encouraging children to venture outside their comfort zone, fathers stimulate a broad spectrum of emotions, thereby expanding children's emotional repertoire, and engaging their developing regulatory abilities  [@flouriFathersBehaviorsChildren2010; @bogelsDoesFatherKnow2011; @parkeFatherInvolvement2000].  
  In line with this theory, empirical research has found that mothers' responses to emotional expression are more warm and supportive than fathers' responses, which are more demanding and punitive [@cassanoInfluenceGenderParental2007; @eisenbergParentalReactionsChildren1999]. In adolescence, maternal warmth was found to be a more salient socialization mechanism of emotional adjustment than maternal demandingness, whereas this pattern was reversed for fathers [@vanlissaRoleFathersMothers2019]. Similar differentiated associations were found for younger children's psychological adjustment [@mcdowellDifferencesMothersFathers2003]. Finally, there is initial evidence that emotional adjustment is more negatively impacted by mothers' negative parenting, compared to fathers [@changHarshParentingRelation2003]; perhaps because negative parenting is somewhat incongruous with the warm "attachment" relationship typically expected of mothers. Based on this research, we therefore hypothesize that emotional adjustment would be more strongly associated with maternal warmth and anger, and with paternal demandingness, operationalized as consistency and inductive reasoning, $H_{3}$.

@@

We have done as requested; the first two paragraphs now introduce the two major goals of the paper, in conceptual terms:

> Children's emotional adjustment is defined as the absence of internalizing symptoms and emotional dysregulation [@goodmanPsychometricPropertiesStrengths2001]. It is integral to social competence, academic achievement, and mental health [@grossHandbookEmotionRegulation2013]. Ample research has demonstrated associations between children's emotional adjustment and parenting; both in terms of quantitative involvement (the amount of time parents spend caring for their child), and in terms of qualitative parenting behaviors (how parents behave towards their child). Nevertheless, the existing literature has two major shortcomings. Firstly, although it is widely acknowledged that quantitative and qualitative involvement are both relevant to child development [@faganShouldResearchersConceptualize2014; @pleckPaternalInvolvementRevised2004], these dimensions are rarely jointly investigated, let alone in interaction [@craigDoesFatherCare2006; @faganShouldResearchersConceptualize2014]. Therefore, we currently do not know how the interplay between quantitative and qualitative involvement relates to emotional development. For example, children’s emotional adjustment might be differently impacted by anger coming from a highly involved parent, than from a generally absent parent. Conversely, children’s poor emotional adjustment might elicit warm responses from hands-on parents, who are aware of the reasons for their children’s emotional troubles, than from out-of-touch parents, who might perceive maladjustment as a problem behavior that should be disciplined with anger or inductive reasoning. The present study overcomes this limitation by investigating both qualitative and quantitative involvement, and the interplay between them, in relation to children's emotional adjustment.  
  Secondly, although emotional adjustment is one particular area of development where fathers have been theorized to play a unique role [@paquetteTheorizingFatherchildRelationship2004], fathers are still not consistently included in research [@faganShouldResearchersConceptualize2014]. Consequently, prior research has not sufficiently addressed mothers' and fathers' joint and unique contributions to emotional adjustment development. The present study overcomes this limitation by incorporating both parents. We use sophisticated methods to examine how joint and relative quantitative involvement at age 6 moderate the longitudinal interplay between mothers' and fathers' qualitative parenting and children's emotional adjustment, from the age of 6 to 12 years. We thus obtain a more comprehensive and nuanced understanding of how fathers' and mothers' involvement relate to children's emotional adjustment development.

@@

We appreciate the suggestion; indeed, the idea that both quantitative and qualitative involvement are relevant for emotional adjustment development (and might interact) is paramount, and the rough division along disciplinary lines is secondary. To address this comment, we have rewritten the argument to focus on the importance of jointly investigating quantitative and qualitative involvement, and removed all references to disciplinary differences.

@@

We have done as requested; the first two paragraphs introduce the study goals, and each subsequent section walks the reader through relevant existing literature, which are then synthesized intospecific labelled hypotheses.

@@

> We therefore hypothesize that parents will display more warmth, consistency, and inductive reasoning towards girls than boys, and more anger towards boys than girls, $H_{7a-d}$. Furthermore, we predict stronger associations between qualitative parenting and emotional adjustment between children and their same-sex parent, $H_{8a-d}$. 
*[...]*
Consequently, we have no hypothesis regarding sex differences in average levels of emotional adjustment.

@@

> Bayes is a full-information estimator, which means that all available information is used without imputing missing values, and consistent estimation is guaranteed under the assumption that data are missing-at-random (a non-parameteric MCAR test resulted in $p > .05$).

@@

> Between-family differences in the interplay between qualitative parenting behaviors and emotional adjustment are estimated as random effects, and we can include between-family level moderators as predictors of these random effects. In this way, we included joint and relative quantitative involvement, __and child sex__ as moderators.

@@

>We examined differences between fathers and mothers in parent- and child effects the same way. [by testing the significance of the difference between corresponding mother- and father effects (see Table 2 and Supplementary Table S5)]. All effects that were significant for one parent but not significant for the other parent had significant difference tests, except for the child effect of emotional adjustment on parental anger, which did not differ significantly between parents.

@@

> Fathers' qualitative involvement is associated with emotional adjustment, even after accounting for mothers [@flouriFathersBehaviorsChildren2010]. Some research suggests that both parents influence children similarly, but that effects might be stronger for mothers [@lambDevelopmentSignificanceFatherChild2004; @faganShouldResearchersConceptualize2014]. Others have argued that mothers and fathers might contribute through somewhat different mechanisms. Attachment relationships are grounded in a balance between warmth and security, and exploration [@grossmannDeveloppementAttachementAdaptation1998]. Some have proposed that warmth is more central to mother-child attachment relationships, whereas father-child relationships can be construed as "activation relationships", with greater emphasis on exploration [@paquetteFatherChildActivationRelationship2013]. By encouraging children to venture outside their comfort zone, fathers stimulate a broad spectrum of emotions, thereby expanding children's emotional repertoire, and engaging their developing regulatory abilities  [@flouriFathersBehaviorsChildren2010; @bogelsDoesFatherKnow2011; @parkeFatherInvolvement2000].  
  In line with this theory, empirical research has found that mothers' responses to emotional expression are more warm and supportive than fathers' responses, which are more demanding and punitive [@cassanoInfluenceGenderParental2007; @eisenbergParentalReactionsChildren1999]. In adolescence, maternal warmth was found to be a more salient socialization mechanism of emotional adjustment than maternal demandingness, whereas this pattern was reversed for fathers [@vanlissaRoleFathersMothers2019]. Similar differentiated associations were found for younger children's psychological adjustment [@mcdowellDifferencesMothersFathers2003]. Finally, there is initial evidence that emotional adjustment is more negatively impacted by mothers' negative parenting, compared to fathers [@changHarshParentingRelation2003]; perhaps because negative parenting is somewhat incongruous with the warm "attachment" relationship typically expected of mothers. Based on this research, we therefore hypothesize that emotional adjustment would be more strongly associated with maternal warmth and anger, and with paternal demandingness, operationalized as consistency and inductive reasoning, $H_{3}$.

@@

> Negative parenting, which we operationalized as parental anger, is a qualitative parenting dimension not originally subsumed in the parenting styles typology. It is negatively linked with children's emotional adjustment [@changHarshParentingRelation2003; @eisenbergParentalReactionsChildren1999]. Parental anger negatively impacts children's emotional adjustment [@scaramellaParentingChildrenDistress2008], because it is upsetting, and because anger reflects poor parental emotion regulation, which serves as a model for children [@changHarshParentingRelation2003]. We thus hypothesized that anger would be negatively related to emotional adjustment, $H_{1d}$.

@@

> [...] bi-directional effects are not routinely investigated in the childhood literature [@flouriFathersBehaviorsChildren2010]. Consequently, most of the research cited in support of hypotheses $H_{1a-d}$ is cross-sectional, and could be taken as support for child effects as well as parent effects. We therefore hypothesize that child effects are similar in direction, and that emotional adjustment positively predicts warmth, consistency, and inductive reasoning, and negatively predicts anger, $H_{2a-d}$. This is supported by studies that investigated child effects specifically, and found that children's maladjustment elicits less parental warmth, and more negative parenting [@gadeyneLongitudinalRelationsParenting2004]. Bi-directional positive links between parental consistency and emotional adjustment have also been reported [@lenguaBidirectionalAssociationsTemperament2005]. 
The adolescent literature provides additional evidence that emotional adjustment positively predicts parental support [@otterpohlCrossLaggedRelationsParenting2015; @vandergiessenPerceivedAutonomySupport2014], and negatively predicts negative parenting [@wijsbroekDirectionEffectsPerceived2011]. Thus, in a way, parents respond positively to emotional adjustment, and negatively to maladjustment.

@@

> we are aware of only one prior study that controlled for stable between-family differences, and similarly reported mostly child effects and between-family level correlations in a sample of adolescents [@vanlissaRoleFathersMothers2019]. The authors explained their findings by proposing that child effects might be more common in adolescence than in childhood, because parent-adolescent relationships are more egalitarian at that age. They argued that the between-family level correlations might reflect the consolidated influence of within-family parent effects that occurred at an earlier age. The fact that the present study similarly found a preponderance of child effects and between-family level correlations in a sample of younger children, age 6-12, suggests that either parent effects occurred at an even younger age, or that something else is at play.

@@

> We found only limited evidence for within-family predictive effects, most of which were child effects on parents. Instead, we mostly found evidence for between-family level correlations. We are not the first to find that parent effects are less prevalent once stable between-family differences are accounted for [@keijsersParentalMonitoringAdolescent2016]. In the area of parenting and emotional adjustment development, we are aware of only one prior study that controlled for stable between-family differences, and similarly reported mostly child effects and between-family level correlations in a sample of adolescents [@vanlissaRoleFathersMothers2019]. The authors explained their findings by proposing that child effects might be more common in adolescence than in childhood, because parent-adolescent relationships are more egalitarian at that age. They argued that the between-family level correlations might reflect the consolidated influence of within-family parent effects that occurred at an earlier age. The fact that the present study similarly found a preponderance of child effects and between-family level correlations in a sample of younger children, age 6-12, suggests that either parent effects occurred at an even younger age, or that something else is at play.  
  As similar patterns continue to emerge, it might be worth reflecting on their implications for developmental theory. Our findings are less congruent with a fully bidirectional conceptualization of reciprocal influence, and more consistent with a model where parents "set the stage" for children's development through their stable differences from other parents, and continue to adjust to children's changing needs over time. Our findings serve as a reminder that we should be critical of causal interpretations of research that does not control for stable between-family differences, not only in the area of emotional adjustment [e.g., @brandPerceivedParentalInconsistency1990], but in developmental research in general. We should reconsider the question: To what extent are parents shaping children's development, and to what extent are they adapting to it?  
  Future research might also devote greater attention to between-family level correlations as an outcome of interest. These correlations can reflect the consolidated effects of parenting effects that occur at an early age [@vanlissaRoleFathersMothers2019], or alternatively, they can reflect the influence of third variables that impact both constructs. We should seek to understand why between-family level correlations are so prominent; for example, by comparing their strength in samples with children of different ages, including very young children, or by examining the influence of more sociological factors. For example, economic hardship undermines parents' capacity for consistent parenting, as well as children's socio-emotional adjustment [@mcloydImpactEconomicHardship1990]. Due to disciplinary differences, such sociological factors are rarely examined in conjunction with sophisticated models of reciprocal influences between parents and children.

@@

> We established direction of effects by testing the difference between corresponding parent- and child effects (see Table 2 and Supplementary Table S5). All effects that were significant in one direction, and non-significant in the opposite direction, had significant difference tests. All effects that were significant (or non-significant) in both directions had non-significant difference tests.

@@

> To address our research questions, we used multilevel, autoregressive structural equation modeling [@asparouhovDynamicStructuralEquation2018]. In this approach, the variance in qualitative parenting behaviors and emotional adjustment is partialized into stable between-individual differences (the mean level for that individual across all measurement occasions), and within-individual fluctuations (deviations from this person-specific mean). This has several advantages: First, controlling for stable between-family differences allows us to estimate the direction of within-family effects more accurately [@hamakerCritiqueCrosslaggedPanel2015]. Any stable factors which might bias cross-lagged effects - such as the influence of SES or neighborhood effects - are accounted for at the between level, and reflected in the correlations between the random intercepts. Secondly, this approach enables us to directly test our hypothesis that between-family differences in quantitative involvement moderate the interplay between qualitative involvement and emotional adjustment: Between-family differences in the interplay between qualitative parenting behaviors and emotional adjustment are estimated as random effects, and we can include between-family level moderators as predictors of these random effects. In this way, we included joint and relative quantitative involvement, and child sex as moderators. Figure 1 shows a graphical representation of the resulting model. We conducted separate analyses for the four qualitative parenting dimensions, because a model including all four variables at once was too complex to estimate.

@@

>  Counter to hypothesis $H_{6a}$, we found that greater joint involvement negatively predicted emotional adjustment. Thus, children in families with highly involved parents generally displayed poorer emotional adjustment. Hypothesis $H_{6b}$ was supported: Children's emotional adjustment was lower in families where mothers were relatively more involved than fathers.

@@

<!--
\section*{References}

\setlength{\leftskip}{0.4in}
\setlength{\parindent}{-0.4in}
-->
